Optimizing Abstract Abstract Machines

### 2012-11-14 paper/paper.tex
Rewrote store-widening section

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\varsigma_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\varsigma_0} : \mathcal{P}(\mathit{State}) \to \mathcal{P}(\mathit{State})
 \text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the inital state:
\begin{equation*}
  F_{\varsigma_0}(S) = \{ \varsigma_0 \} \cup \{ \varsigma' \mathrel{|} \varsigma \in S \text{ and } \varsigma \machstep \varsigma' \}\text. 
\end{equation*}
Then, the evaluator for the analysis computes the least fixed point of the global transfer function:
\begin{equation*}
 \eval(e) = \mathrm{lfp}(F_{\varsigma_0})\text{,}
\end{equation*}
where:
\begin{equation*}
\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\text.
\end{equation*}

@@

A common technique to accelerate convergence in flow analyses is to share
a common store that is monotonically updated as interpretation
continues. In the ISWIM language with only unary operators, this makes
our 0-CFA quartic in the size of the program, which in one step drops the
complexity from intractable exponentiality to merely daunting
polynomiality.

We can reuse the reduction relation we have defined, but with some
scaffolding around it. A state in the store-widened machine is a pair
of a set of ISWIM states without a store, and a store. A step of this
machine will pair the store with each of the states in this set,
reduce, and combine the output stores to create the next. The
scaffolded reduction relation, $\widehat{\machstep}$, is as follows:

\begin{align*}
(cs, \msto) &\mathrel{\widehat{\machstep}} (cs', \msto') \\
\mbox{where } cs' &= \{c' \mid wn(c, \msto) \mathrel{\machstep} wn(c, \msto^c), c \in cs\} \\
              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\mathrel{\machstep} wn(c', \msto^c), c \in cs\} \\
wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
That is, the next set of states is what we get by stepping each
given state, without its store. The next store is the combination of all produced
stores. For simplicity, assume that stuck states step to themselves.

>>>>

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\varsigma_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\varsigma_0} : \mathcal{P}(\mathit{State}) \to \mathcal{P}(\mathit{State})
 \text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the inital state:
\begin{equation*}
  F_{\varsigma_0}(S) = \{ \varsigma_0 \} \cup \{ \varsigma' \mathrel{|} \varsigma \in S \text{ and } \varsigma \machstep \varsigma' \}\text. 
\end{equation*}
Then, the evaluator for the analysis computes the least fixed point of the global transfer function:
\begin{equation*}
 \eval(e) = \mathrm{lfp}(F_{\varsigma_0})\text{,}
\end{equation*}
where $\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}$.

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0CFA quartic in the size of the
program.
%
Thus, in one step, complexity drops from intractable exponentiality to merely
daunting polynomiality.

Since we can cast this optimization as a widening, there is no need to change
the transition relation itself.
%
Rather, what changes is the structure of the fixed-point iteration.
%
In each pass, the algorithm will collect all newly produced stores and join
them together.
%
Then, before each transition, it will install this joined store into current
state.

To describe this process, we'll refactor the transition relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The refactored relation, $\widehat{\machstep}$, becomes:
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid wn(c, \msto) \mathrel{\machstep} wn(c, \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid wn(c,\msto)\mathrel{\machstep} wn(c', \msto^c), c \in C\} \\
wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all subsequent stores.
